---
title: "Using the QuakeCoursera Package"
author: "Jake Markman"
date: "7/11/2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the QuakeCoursera Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE, messages = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(QuakeCursera)
library(ggplot2)
library(dplyr)
library(readr)
library(leaflet)
```  

QuakeCoursera is a package developed as the capstone project for the Advanced 
R Programming specialization on [Coursera](http://www.coursera.org).  It exists
to help clean and visualize data from the [NOAA Significant Earthquake Database](https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1)  
  
This document serves as an introduction to the package and demonstrates
how to use each of the individual features.

This document is organized into three broad categores, as follows:  

1.  [Getting and Cleaning the Data](data)  
2.  [Timeline Visualization](timeline)  
3.  [Interactive Mapping](mapping)  

## Getting and Cleaning the Data  
<a id="data"></a>

The [NOAA Significant Earthquake Database](https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1) 
is an amazing resource.  It has information on earthquakes that have occurred worldwide over more than 
4000 years.  However as a raw text file, it isn't the easist to work with.  

Some cleaning is required to get the data ready to work nicely with the visualization
functions provided below.  The easist thing to do is simply to load the clean dataset 
included with this package using the `data()` function:  

```{r}
data('clean_NOAA')
```  

The drawback with this approach is that the included dataset has been frozen as of the
date of the download for creation of this package, July 1 2017.  You can download the 
most recent dataset yourself [here](https://www.ngdc.noaa.gov/nndc/struts/results?type_0=Exact&query_0=$ID&t=101650&s=13&d=189&dfn=signif.txt) 
and clean it using the `eq_clean_data()` function.  Assuming you have downloaded 
the raw data to your working directory as 'signif.txt':  

```{r eval = FALSE}
clean_NOAA <- readr::read_delim('signif.txt',delim = '\t') %>%
    eq_clean_data()
```  

Eather way, `clean_NOAA` is now ready to be used for visualizing the historical 
earthquake data.  

### Unexported Helper Functions  

To minimize 



<a id="timeline"></a>
<a id="mapping"></a>
